\documentclass[
]{jss}

%% recommended packages
\usepackage{orcidlink,thumbpdf,lmodern}

\usepackage[utf8]{inputenc}

\author{
Piotr Chlebicki\\Stockholm University \And Maciej
Beręsewicz~\orcidlink{0000-0002-8281-4301}\\Poznań University of
Economics\\
and Business
}
\title{Single-Source Capture-Recapture Models With \pkg{singleRcapture}}

\Plainauthor{Piotr Chlebicki, Maciej Beręsewicz}
\Plaintitle{Single-Source Capture-Recapture Models With singleRcapture}
\Shorttitle{\pkg{singleRcapture}: Single-Source Capture-Recapture
Models}


\Abstract{
Estimating population size is an important issue in official statistics,
social sciences and natural sciences. One way to approach this problem
is to use capture-recapture methods, which can be classified according
to the number of sources used, the main distinction being between
methods based on one source and those based on two or more sources. In
this presentation we will introduce the \pkg{singleRcapture} R package
for fitting SSCR models. The package implements state-of-the-art models
as well as some new models proposed by the authors (e.g.~extensions of
zero-truncated one-inflated and one-inflated zero-truncated models). The
software is intended for users interested in estimating the size of
populations, particularly those that are difficult to reach or for which
information is available from only one source and dual/multiple system
estimation cannot be used.
}

\Keywords{population size estimation, truncated distributuons, count
regression models, \proglang{R}}
\Plainkeywords{population size estimation, truncated
distributuons, count regression models, R}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
    Piotr Chlebicki\\
    Stockholm University\\
    Matematiska institutionen\\
106 91 Stockholm, Albano hus 1\\
  E-mail: \email{piotr.chlebicki@math.su.se}\\
  URL: \url{https://github.com/Kertoo}\\~\\
      Maciej Beręsewicz\\
    Poznań University of Economics\\
and Business\\
    Poznań University of Economics and Business\\
Department of Statistics\\
Institute of Informatics and Quantitative Economics\\
Al. Niepodległosci 10\\
61-875 Poznań, Poland\\
  E-mail: \email{maciej.beresewicz@ue.poznan.pl}\\
  
  }


% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}




\usepackage{amsmath, amsthm, amssymb} \usepackage{calc, ragged2e}

\DeclareMathAlphabet{\mathmybb}{U}{bbold}{m}{n}
\newcommand{\1}{\mathcal{I}} \newcommand{\bx}{\boldsymbol{x}}
\newcommand{\bX}{\boldsymbol{X}} \newcommand{\bbeta}{\boldsymbol{\beta}}
\newcommand{\boeta}{\boldsymbol{\eta}}

\begin{document}



\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\allowdisplaybreaks

\hypertarget{literature-review}{%
\subsection{Literature review}\label{literature-review}}

This work is supported by the National Science Center, OPUS 20 grant no.
2020-/-39/-B/-HS4/-00941
\textit{Towards census-like statistics for foreign-born populations -- quality, data integration and estimation}

The subject of this workshop is the \pkg{singleRcapture} package and its
lightweight extension that allows for integration with other
\proglang{R} packages called \pkg{singleRcaptureExtra}.

The package is available on CRAN:
\url{CRAN.R-project.org/package=singleRcapture} while the extension is
available on:
\url{https://github.com/ncn-foreigners/singleRcaptureExtra}.

The \pkg{singleRcapture} package is an \proglang{R} language package
that focuses on implementing state of the art methods for frequentist
point and interval estimation of size of closed populations in
single-source capture-recapture (SSCR) setting (e.g.~estimation of the
population size of irregular migrants at set time point in a given
area).

The beginning of inference in single source capture-recapture dates back
to the seminal \cite{ztpoisson} paper in which the zero truncated
poisson model was applied to study the size of population of irregular
migrants in fours cities in Netherlands.

\hypertarget{how-do-we-estimate-population-size-with-only-one-register}{%
\subsection{How do we estimate population size with only one
register?}\label{how-do-we-estimate-population-size-with-only-one-register}}

Let \(Y_{k}\) represent the number of times \(k\)-th unit was observed
in source data. Clearly, we don not know how often \(Y_{k}=0\) and to
find the total population size \(N\) we need to estimate it. In general,
we assume that conditional distribution of \(Y_{k}\) given
a\textasciitilde vector of covariates \(\mathbf{x}_{k}\) follows some
version of zero truncated count data distribution. Knowing the
parameters of the distribution we may estimate the population size using
Horwitz-Thompson type estimator: \begin{equation*}
    \hat{N}=
    \sum_{k=1}^{N}\frac{I_{k}}{\mathbb{P}[Y_{k}>0|\mathbf{x}_{k}]}=
    \sum_{k=1}^{N_{obs}}\frac{1}{\mathbb{P}[Y_{k}>0|\mathbf{x}_{k}]},
\end{equation*} where \(I_{k}:=\mathcal{I}_{\mathbb{N}}(Y_{k})\), and
maximum likelihood estimate of \(N\) is obtained after substituting
regression estimates for \(\mathbb{P}[Y_{k}>0|\mathbf{x}_{k}]\) into the
equation above. Most of the methods relate to poisson processes.

The analytic variance estimation is then done by computing two parts of
the decomposition due to the law of total variance: \begin{equation*}
  \text{var}[\hat{N}]=
  \mathbb{E}\left[\text{var}
  \left[\hat{N}|I_{1},\dots,I_{n}\right]\right]+
  \text{var}\left[\mathbb{E}[\hat{N}|I_{1},\dots,I_{n}]\right]
\end{equation*} where the first addend is by the multivariate \(\delta\)
method seen to be: \begin{equation*}
  \mathbb{E}\left[\text{var}
  \left[\hat{N}|I_{1},\dots,I_{n}\right]\right]=
  \left.\left(\frac{\partial(N|I_1,\dots,I_N)}{\partial\boldsymbol{\beta}}\right)^{T}
  \text{cov}\left[\boldsymbol{\beta}\right]
  \left(\frac{\partial(N|I_1,\dots,I_N)}{\partial\boldsymbol{\beta}}\right)
  \right|_{\boldsymbol{\beta}=\hat{\boldsymbol{\beta}}}
\end{equation*} while the later part of the decomposition is under the
assumption of independence of \(I_{k}\)'s optimally estimated via:
\begin{align}
  \text{var}\left(\mathbb{E}(\hat{N}|I_{1},\dots,I_{n})\right)&=
  \text{var}\left(\sum_{k=1}^{N}\frac{I_{k}}{\mathbb{P}(Y_{k}>0)}\right)\nonumber\\
  &=\sum_{k=1}^{N}\text{var}\left(\frac{I_{k}}{\mathbb{P}(Y_{k}>0)}\right)\nonumber\\
  &=\sum_{k=1}^{N}\frac{1}{\mathbb{P}(Y_{k}>0)^{2}}\text{var}(I_{k})\nonumber\\
  &=\sum_{k=1}^{N}\frac{1}{\mathbb{P}(Y_{k}>0)^{2}}\mathbb{P}(Y_{k}>0)(1-\mathbb{P}(Y_{k}>0))\nonumber\\
  &=\sum_{k=1}^{N}\frac{1}{\mathbb{P}(Y_{k}>0)}(1-\mathbb{P}(Y_{k}>0))\nonumber\\
  &\approx\sum_{k=1}^{N}\frac{I_{k}}{\mathbb{P}(Y_{k}>0)^{2}}(1-\mathbb{P}(Y_{k}>0))\nonumber\\
  &=\sum_{k=1}^{N_{obs}}\frac{1-\mathbb{P}(Y_{k}>0)}{\mathbb{P}(Y_{k}>0)^{2}}
\end{align}

\subsection[Example with R code]{Example with \proglang{R}
code}\label{r-code}

Installation:

\begin{CodeChunk}
\begin{CodeInput}
R> install.packages("singleRcapture")
R> remotes::install_github(
+   "https://github.com/ncn-foreigners/singleRcaptureExtra"
+ )
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> library(singleRcapture)
R> 
R> head(netherlandsimmigrant)
\end{CodeInput}
\begin{CodeOutput}
  capture gender    age       reason       nation
1       1   male <40yrs Other reason North Africa
2       1   male <40yrs Other reason North Africa
3       1   male <40yrs Other reason North Africa
4       1   male <40yrs Other reason         Asia
5       1   male <40yrs Other reason         Asia
6       2   male <40yrs Other reason North Africa
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> summary(netherlandsimmigrant)
\end{CodeInput}
\begin{CodeOutput}
    capture         gender         age                reason    
 Min.   :1.000   female: 398   <40yrs:1769   Illegal stay: 259  
 1st Qu.:1.000   male  :1482   >40yrs: 111   Other reason:1621  
 Median :1.000                                                  
 Mean   :1.162                                                  
 3rd Qu.:1.000                                                  
 Max.   :6.000                                                  
                    nation    
 American and Australia: 173  
 Asia                  : 284  
 North Africa          :1023  
 Rest of Africa        : 243  
 Surinam               :  64  
 Turkey                :  93  
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> basicModel <- estimatePopsize(
+   formula = capture ~ gender + age + nation,
+   model   = ztpoisson(),
+   data    = netherlandsimmigrant
+ )
R> 
R> summary(basicModel)
\end{CodeInput}
\begin{CodeOutput}

Call:
estimatePopsize.default(formula = capture ~ gender + age + nation, 
    data = netherlandsimmigrant, model = ztpoisson())

Pearson Residuals:
     Min.   1st Qu.    Median      Mean   3rd Qu.      Max. 
-0.486442 -0.486442 -0.298080  0.002093 -0.209444 13.910844 

Coefficients:
-----------------------
For linear predictors associated with: lambda 
                     Estimate Std. Error z value  P(>|z|)    
(Intercept)           -1.3411     0.2149  -6.241 4.35e-10 ***
gendermale             0.3972     0.1630   2.436 0.014832 *  
age>40yrs             -0.9746     0.4082  -2.387 0.016972 *  
nationAsia            -1.0926     0.3016  -3.622 0.000292 ***
nationNorth Africa     0.1900     0.1940   0.979 0.327398    
nationRest of Africa  -0.9106     0.3008  -3.027 0.002468 ** 
nationSurinam         -2.3364     1.0136  -2.305 0.021159 *  
nationTurkey          -1.6754     0.6028  -2.779 0.005445 ** 
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

AIC: 1712.901
BIC: 1757.213
Residual deviance: 1128.553

Log-likelihood: -848.4504 on 1872 Degrees of freedom 
Number of iterations: 8
-----------------------
Population size estimation results: 
Point estimate 12690.35
Observed proportion: 14.8% (N obs = 1880)
Std. Error 2808.169
95% CI for the population size:
          lowerBound upperBound
normal      7186.444   18194.26
logNormal   8431.275   19718.32
95% CI for the share of observed population:
          lowerBound upperBound
normal     10.332927   26.16037
logNormal   9.534281   22.29793
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{detailed-information}{%
\section{Detailed information}\label{detailed-information}}

\hypertarget{fitting-method}{%
\subsection{Fitting method}\label{fitting-method}}

As previously showcased the \pkg{singleRcapture} package supports
modelling (linear) dependence on covariates of all parameters. To that
end a modified IRLS algorithm is employed, full details are available in
\cite{VGAM-main}. In order to employ the algorithm a modified model
matrix is created \(\boldsymbol{X}_{\text{vlm}}\) at call to
\texttt{estimatePopsize}. In the context of the models implemented in
\pkg{singleRcapture} this matrix can be writen as:

\begin{equation}
  \boldsymbol{X}_{vlm}=
  \begin{pmatrix}
    \boldsymbol{X}_{1} & \boldsymbol{0} &\dotso &\boldsymbol{0}\cr
    \boldsymbol{0} & \boldsymbol{X}_{2} &\dotso &\boldsymbol{0}\cr
    \vdots & \vdots & \ddots & \vdots\cr
    \boldsymbol{0} & \boldsymbol{0} &\dotso &\boldsymbol{X}_{p}
  \end{pmatrix}
\end{equation}

In the context of multi-parameter families we have a matrix of linear
predictors \(\boldsymbol{\eta}\) instead of a vector, with the number of
columns matching the number of parameters in the distribution.
``Weights'' are then modified to be information matrices
\(\displaystyle\mathbb{E}\left[-\frac{\partial^{2}\ell}{\partial\boldsymbol{\eta}_{(k)}^{T}\partial\boldsymbol{\eta}_{(k)}}\right]\)
where \(\boldsymbol{\eta}_{(k)}\) is the \(k\)'th row of
\(\boldsymbol{\eta}\), while in the usual IRLS they are scalars
\(\displaystyle\mathbb{E}\left[-\frac{\partial^{2}\ell}{\partial\eta_{k}^{2}}\right]\)
which is often just
\(\displaystyle-\frac{\partial^{2}\ell}{\partial\eta^{2}}\).

\begin{enumerate}
    \justifying
    \item Initialize with $\texttt{converged} \leftarrow$ FALSE, $\texttt{iter}\leftarrow 1, \boldsymbol{\eta}\leftarrow\texttt{start}, \boldsymbol{W}\leftarrow I, \ell\leftarrow\ell(\boldsymbol{\beta})$.
    \item Store values from the previous step: $\ell_{-}\leftarrow\ell, \boldsymbol{W}_{-}\leftarrow\boldsymbol{W}, \boldsymbol{\beta}_{-}\leftarrow\boldsymbol{\beta}$ (the last assignment is omitted during the first iteration), and assign values in current iteration $\displaystyle\boldsymbol{\eta}\leftarrow\boldsymbol{X}_{\text{vlm}}\boldsymbol{\beta}+\boldsymbol{o}, \boldsymbol{W}_{(k)}\leftarrow\mathbb{E}\left[-\frac{\partial^{2}\ell}{\partial\boldsymbol{\eta}_{(k)}^{T}\partial\boldsymbol{\eta}_{(k)}}\right], Z\leftarrow\boldsymbol{\eta}_{(k)}+\frac{\partial\ell}{\partial\boldsymbol{\eta}_{(k)}}\boldsymbol{W}_{(k)}^{-1}-\boldsymbol{o}_{(k)}$.
    \item Assign current coefficient value: $\boldsymbol{\beta}\leftarrow\left(\boldsymbol{X}_{\text{vlm}}\boldsymbol{W}\boldsymbol{X}_{\text{vlm}}\right)^{-1}\boldsymbol{X}_{\text{vlm}}\boldsymbol{W}\boldsymbol{Z}$.
    \item If $\ell(\boldsymbol{\beta})<\ell(\boldsymbol{\beta}_{-})$ try selecting the smallest value $h$ such that for $\boldsymbol{\beta}_{h}\leftarrow2^{-h}\left(\boldsymbol{\beta}+\boldsymbol{\beta}_{-}\right)$ the inequality $\ell(\boldsymbol{\beta}_{h})>\ell(\boldsymbol{\beta}_{-})$ holds if this is successful $\boldsymbol{\beta}\leftarrow\boldsymbol{\beta}_{h}$ else stop the algorithm.
    \item If convergence is achieved end algorithm, else return to step 2.
\end{enumerate}

\hypertarget{avaiable-models}{%
\subsection{Avaiable models}\label{avaiable-models}}

The full list of implemented models in \pkg{singleRcapture} along with
the expressions for probability density functions and point estimates is
found in the collective help file for all family functions:

\begin{CodeChunk}
\begin{CodeInput}
R> ?ztpoisson
\end{CodeInput}
\end{CodeChunk}

Here we limit ourselves to just listing the family functions:

\begin{itemize}
    \item Zero-truncated and zero-one-truncated Poisson, geometric, NB type II regression where the untruncated distribution is parameterized as:
    \begin{equation*}
        \mathbb{P}[Y=y|\lambda,\alpha] = \frac{\Gamma\left(y+\alpha^{-1}\right)}{\Gamma\left(\alpha^{-1}\right)y!}
        \left(\frac{\alpha^{-1}}{\alpha^{-1}+\lambda}\right)^{\alpha^{-1}}
        \left(\frac{\lambda}{\lambda + \alpha^{-1}}\right)^{y}.
    \end{equation*}
    \item Zero-truncated one-inflated (ztoi) modifications distributions where the new probability $\mathbb{P}^{\ast}$ measure is defined in terms of count data measure $\mathbb{P}$ with support on $\mathbb{N}\cup\{0\}$ as:
    \begin{equation*}
    \mathbb{P}^{\ast}[Y=y]=
    \begin{cases}
    \mathbb{P}[Y=0] & y=0, \\
    \omega\left(1-\mathbb{P}[Y=0]\right)+(1-\omega)\mathbb{P}[Y=1] & y=1, \\
    (1-\omega)\mathbb{P}[Y=y] & y>1,
    \end{cases}
    \end{equation*}
    \begin{equation*}
        \mathbb{P}^{\ast}[Y=y|Y>0]=\omega\mathcal{I}_{\{1\}}(y)+(1-\omega)\mathbb{P}[Y=y|Y>0].
    \end{equation*}
    \item One-inflated zero-truncated (oizt) modifications distributions where the new probability $\mathbb{P}^{\ast}$ measure is defined as:
    \begin{equation*}
        \mathbb{P}^{\ast}[Y=y] = \omega \mathcal{I}_{\{1\}}(y)+(1-\omega)\mathbb{P}[Y=y],
    \end{equation*}
    \begin{equation*}
        \mathbb{P}^{\ast}[Y=y|Y>0] = 
        \omega\frac{\mathcal{I}_{\{1\}}(y)}{1-(1-\omega)\mathbb{P}[Y=0]}+
        (1-\omega)\frac{\mathbb{P}[Y=y]}{1-(1-\omega)\mathbb{P}[Y=0]}.
    \end{equation*}
    \item Generalized Chao's and Zelterman's estimators via logistic regression on variable $Z$ defined as $Z=1$ if $Y=2$ and $Z=0$ if $Y=1$ with $Z\sim b(p)$ where $\text{logit}(p)=\ln(\lambda/2)$ for poisson parameter $\lambda$,
    \begin{align}
        \hat{N} &= N_{obs}+
        \sum_{k=1}^{\boldsymbol{f}_{1}+\boldsymbol{f}_{2}}
        \left(2\exp\left(\boldsymbol{x}_{k}\hat{\boldsymbol{\beta}}\right)+
        2\exp\left(2\boldsymbol{x}_{k}\hat{\boldsymbol{\beta}}\right)\right)^{-1},
        \tag{\text{Chao's estimator}}\\
        \hat{N}&=\sum_{k=1}^{N_{obs}}
        \left(1-\exp\left(-2\exp\left(\boldsymbol{x}_{k}\hat{\boldsymbol{\beta}}\right)\right)\right)^{-1}.
        \tag{\text{Zelterman's estimator}}
    \end{align}
    \item Alternative approaches to modelling one-inflation that mimic hurdle models where the first type zero truncated hurdle model (ztHurdle) is defined as:
    \begin{equation*}
        \mathbb{P}^{\ast}[Y=y]=\begin{cases}
        \frac{\mathbb{P}[Y=0]}{1-\mathbb{P}[Y=1]} & y=0, \\
        \pi(1-\mathbb{P}[Y=1]) & y=1, \\
        (1-\pi) \frac{\mathbb{P}[Y=y]}{1-\mathbb{P}[Y=1]} & y>1,
        \end{cases}
    \end{equation*}
    \begin{equation*}
        \mathbb{P}^{\ast}[Y=y|Y>0]=\pi\mathcal{I}_{\{1\}}(y)+
        (1-\pi)\mathcal{I}_{\mathbb{N}\setminus\{1\}}(y)\frac{\mathbb{P}[Y=y]}{1-\mathbb{P}[Y=0]-\mathbb{P}[Y=1]}
    \end{equation*}
    \item The Hurdle zero truncarted (Hurdlezt) is defined as:
    \begin{align*}
        \mathbb{P}^{\ast}[Y=y]&=\begin{cases}
        \pi & y=1, \\
        (1-\pi) \frac{\mathbb{P}[Y=y]}{1-\mathbb{P}[Y=1]} & y\neq1,
        \end{cases}\\
        \mathbb{P}^{\ast}[Y=y|Y>0]&=\begin{cases}
            \pi\frac{1-\mathbb{P}[Y=1]}{1-\mathbb{P}[Y=0]-\mathbb{P}[Y=1]} & y=1,\\
            (1-\pi)\frac{\mathbb{P}[Y=y]}{1-\mathbb{P}[Y=0]-\mathbb{P}[Y=1]} & y>1.
        \end{cases}
    \end{align*}
\end{itemize}

\hypertarget{structure-of-a-family-function}{%
\subsection{Structure of a family
function}\label{structure-of-a-family-function}}

\itemize{
  \item \code{makeMinusLogLike} -- A factory function for creating the:
  \begin{equation*}
    \ell(\boldsymbol{\beta}), 
    \frac{\partial\ell}{\partial\boldsymbol{\beta}},
    \frac{\partial^{2}\ell}{\partial\boldsymbol{\beta}^{T}\partial\boldsymbol{\beta}}
  \end{equation*}
  functions from $\boldsymbol{y}$ vector and $\boldsymbol{X}_{vlm}$ the argument \code{deriv} with possible 
  values in \code{c(0, 1, 2)} provides which derivative to return with the default \code{0} being just the minus log-likelihood.
  \item \code{links} -- List with link functions.
  \item \code{mu.eta, variance} -- Functions of linear predictors that return expected value and variance. There is a `type` argument with 2 possible values \code{"trunc"} and \code{"nontrunc"} that specifies whether to return $\mathbb{E}[Y|Y>0], \text{var}[Y|Y>0]$ or $\mathbb{E}[Y], \text{var}[Y]$ respectively, also the \code{deriv} argument with values in \code{c(0, 1, 2)} is used for indicating the derivative with respect to the linear predictors with is used for providing standard error in \code{predict} method.
  \item \code{family} -- Character that specifies name of the model.
  \item \code{valideta, validmu} -- For now only returns true. In near future will be used to check whether applied linear predictors are valid (i.e. are transformed into some elements of parameter space the subjected to inverse link function).
  \item \code{funcZ, Wfun} -- Functions that create pseudo residuals and working weights used in IRLS algorithm.
  \item \code{devResids} -- Function that given the linear predictors prior weights vector and response vector returns deviance residuals.
  \item \code{pointEst, popVar} -- Functions that given prior weights linear predictors and in the later case also estimation of  $\text{cov}(\hat{\boldsymbol{\beta}})$ and $\boldsymbol{X_{vlm}}$ matrix return point estimate for population size and analytic estimation of its variance.There is a additional boolean parameter \code{contr} in the former function that if set to true returns contribution of each unit.
  \item \code{etaNames} -- Names of linear predictors.
  \item \code{densityFunction} -- A function that given linear predictors returns value of PMF at values \code{x}. Additional argument \code{type} specifies whether to return $\mathbb{P}[Y|Y>0]$ or $\mathbb{P}[Y]$.
  \item \code{simulate} -- A function that generates values of dependent  vector given linear predictors.
  \item \code{getStart} -- Expression for generating starting points.
}

\hypertarget{marginal-frequencies}{%
\subsection{Marginal frequencies}\label{marginal-frequencies}}

A popular method of testing the model fit in single source
capture-recapture studies is comparing the fitted marginal frequencies
\(\displaystyle\sum_{j=1}^{N_{obs}}\hat{\mathbb{P}}\left[Y_{j}=k|\boldsymbol{x}_{j}, Y_{j} > 0\right]\)
with the observed marginal frequencies
\(\displaystyle\sum_{j=1}^{N}\mathcal{I}_{\{k\}}(Y_{k})=\sum_{j=1}^{N_{obs}}\mathcal{I}_{\{k\}}(Y_{k})\)
for \(k\geq1\). If a fitted model bears sufficient resemblance to the
real data collection process these quantities should be quite close and
both \(G\) and \(\chi^{2}\) tests may be employed in order to test the
statistical significance of the discrepancy with the following
\pkg{singleRcapture} syntax:

\begin{CodeChunk}
\begin{CodeInput}
R> (margFreq <- marginalFreq(basicModel))
\end{CodeInput}
\begin{CodeOutput}
$table
           0            1            2            3            4            5 
1.081035e+04 1.612589e+03 2.337193e+02 3.013473e+01 3.242543e+00 2.908758e-01 
           6 
2.214773e-02 

$y
y
   1    2    3    4    5    6 
1645  183   37   13    1    1 

$df
[1] -3

$name
[1] "ztpoisson"

attr(,"class")
[1] "singleRmargin"
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> summary(margFreq, df = 1, dropl5 = "group")
\end{CodeInput}
\begin{CodeOutput}
Test for Goodness of fit of a regression model:

                 Test statistics df P(>X^2)
Chi-squared test           50.06  1 1.5e-12
G-test                     34.31  1 4.7e-09

-------------------------------------------------------------- 
Cells with fitted frequencies of < 5 have been grouped 
Names of cells used in calculating test(s) statistic: 1 2 3  
\end{CodeOutput}
\end{CodeChunk}

where the \texttt{dropl5} argument is used to indicate how to handle the
cells with less than \(5\) fitted observations, note however that
currently there is no continuity correction.

\newpage

\bibliography{refs.bib}



\end{document}
